---
title: "The Role of Shelters in Providing Emergency Housing especially for the Homeless"
author: Syed Hassan
thanks: "Code and data are available at: https://github.com/saiyedgh/Shelters_Toronto.git"
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
abstract: "The city and the people of Toronto provide shelter to thousands of people every year. The majority that takes benefit from the city's overnight shelter are the homeless. The occupancy of these shelters is 100% on most nights and rarely does space go unoccupied. We need to make these shelters more accessible for the homeless and provide more funding to expand the program."
format: pdf
toc: TRUE
number-sections: true
bibliography: references.bib
---

```{r}
#| messages: false
#| echo: false
#| warning: false
#| output: false
#### Preamble ####
# Title: 
# Author: Syed Hassan
# Email: syed.ext@gmail.com
# Date: 16 March 2023
# GitHub: https://github.com/saiyedgh/Shelters_Toronto.git
```


```{r}
#| include: false
#| echo: false
#| warning: false
#| message: false

##### Install and run the required libraries and packages #####

library(opendatatoronto)
library(tidyverse)
library(here)
library(readr)
library(readxl)
library(kableExtra)
library(scales)
library(knitr)
library(janitor)
```
\


```{r}

package <- show_package("21c83b32-d5a8-4106-a54f-010dbe49f6f2")
package

# get all resources for this package
resources <- list_package_resources("21c83b32-d5a8-4106-a54f-010dbe49f6f2")

# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
head(datastore_resources)

# load the 2022 datastore resource as a sample
data_2022 <- filter(datastore_resources, row_number()==6) %>% get_resource()
data_2022
head(data_2022)

data_2021 <- filter(datastore_resources, row_number()==4) %>% get_resource()
data_2021

head(data_2021)

```




# Introduction

Homelessness is a genuine concern of the city and the people of Toronto. The city has acquired plenty of permanent and temporary shelters to provide better housing and overnight shelters for the homeless as well as others seeking emergency housing. The city of Toronto, like many other metropolitan cities, has suffered from the problem of homelessness for years. However, the role of the shelters in helping the homeless might go unnoticed because of the increasing numbers on the streets. My analysis aims to discover the issue in detail using data provided by the city of Toronto. Through my analysis, I aim to take you deep into statistics as well as focus on the patterns in order to develop empathy for the homeless. Lastly, the analysis will conclude the role of the shelters of Toronto in helping people overall and how they can improve to include more people in need.


# Data

Coming soon...

# Results

Coming soon...


# Discussion

Coming soon...


# Limitations

Coming soon...


# Conclusion

Coming soon...


\newpage

# Appendix


\newpage

## Code


\newpage

# References
